---
title: "ENVS-193DS_Winter-2026_final"
author: Henderson Vo
date: 2026-03-17
output: html_document
---
[GitHub repo](https://github.com/hendersonvo/ENVS-193DS_winter-2026_final)

# Set up

```{r message=FALSE, warning=FALSE}
# loading libraries
library(tidyverse)
library(janitor)
library(here)
library(sysfonts)
library(showtext)

# loading data
kelp <- read_csv(here("data", "Annual_Kelp_All_Years_20250903.csv"))
mp_selection <- read_csv(here("data", "MOPL_nest-site_selection_Duchardt_et_al._2019.csv"))

# cleaning data

mp_clean <- mp_selection |> 
  clean_names() 
```



## Problem 1. Research writing

### a. Transparent statistical methods

In part 1, they used a linear regression. In part 2, they used an ANOVA test. 

### b. Figure needed

One figure that could accompany the statement in part 2 is a side-by-side boxplot of wetland area across different water year classifications. We would include water year type on the x-axis and wetland area (in m^2^) on the y-axis. For each year type, we would include a vertically-oriented boxplot with the box representing the 1st quartile, median, and 3rd quartile, and the tails representing 1.5 $\times$ IQR. We would also include underlying data points (observations of wetland area) behind the boxes as a jitter plot. This way, it is possible to visually determine the central tendency and relative variance of each treatment group (year type), which should support the fact that not one treatment group median is significantly different from the others.

### c. Suggestions for rewriting

We have determined that flooded wetland area in the San Joaquin River Delta is not predicted by precipitation (linear regression, F(df) = test statistic, p = 0.11, $\alpha$ = 0.05). We have also found that median flooded wetland area does not differ across water year classification (wet, above normal, below normal, dry, critical drought) (ANOVA, F(df) = test statistic, p = 0.12, $\alpha$ = 0.05).

### d. Interpretation

One additional variable that could influence the flooded area of a wetland is the location of the land parcel, particularly if land parcels are private or public land. This variable is a binary categorical variable (private/not private or public/not public). This variable might influence wetland flooding because management strategies (involving water diversion restriction meant to retain flooded wetlands) are suggested to differ between public and private land parcels, making it plausible that there could be a difference in wetland area between lands that choose to divert water and those that choose to retain water. 


## Problem 2. Data visualization

### a. Cleaning and summarizing

```{r}
kelp_clean <- kelp |> # use kelp data
  clean_names() |> # change column names to lowercase
  
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # filter out observations with this date, transect number, and quad number
  
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # only keep observations with IVEE, NAPL, or CARP as their site
  
  mutate(site_name = case_when(site == "NAPL" ~ "Naples", # create a new column site_name with full names of site codes
                               site == "IVEE" ~ "Isla Vista",
                               site == "CARP" ~ "Carpinteria")) |> 
  
  mutate(site_name = as_factor(site_name), # set site_name to factor instead of string
       site_name = fct_relevel(site_name, # set levels to site_name factors
                               "Naples", "Isla Vista", "Carpinteria")) |> 
  
  group_by(site_name, year) |> # group by site_name and year
  
  summarize(mean_fronds = mean(fronds)) |> # give use the mean frond count
  
  ungroup() # stop grouping by site_name and year

slice_sample(kelp_clean, n = 5)
str(kelp_clean)

```


### b. Understanding the data

Because we are looking at frond observations, we want to exclude observations from 2000-09-22 in transect 6 in quad 40 because they are marked to have -99999 fronds. This does not biologically make sense, but it is instead intended to represent that there is missing data (NA), as indicated by the notes section saying that the datasheet was lost; this reasoning is also outlined in the Methods and Protocols section for this dataset.  

### c. Visualize the data

```{r}
font_add_google("Lato", "lato")
showtext_auto()

ggplot(data = kelp_clean,
       aes(x = year,
           y = mean_fronds,
           color = site_name)) +
  geom_line() +
  geom_point() +
  labs(x = "Year",
       y = "Mean kelp frounds per year",
       title = "Sites vary in mean kelp fronds per year",
       color = "Site") +
  scale_color_manual(values = c("Naples" = "turquoise4",
                                "Isla Vista" = "tan",
                                "Carpinteria" = "darkgrey")) +
  theme_minimal(base_family = "lato") +
  theme(plot.title = element_text(hjust = 0.5))
```


### d. Interpretation

We can see that the site with the most variable mean kelp fronds per year is Isla Vista, which we can visually determine based on how the tan points alternate between being high (higher than the other two sites) and low (about as low as the other sites) on the y-axis. The site with the least variable mean kelp fronds per year is Carpinteria, which we can similarly visually determine based on how the points do not seem to fluctuate in position on the y-axis as much as the other two plots, with the higher points being about as high or lower than Naples (the second-least variable site) and the lower points being about as low or a little higher than the other two sites; inferring that this site has a smaller range of mean kelp frond measurements, which could also infer less variation.  

## Problem 3. Data analysis

### a. Response Variable

### b. Purpose of study

### c. Table of models

### d. Run the models

### e. Select the best model

### f. Check the diagnostics

### g. Visualize the model predictions

### h. Write a caption for your figure

### i. Calculate model predictions

### j. Interpret your results

## Problem 4. Affective and exploratory visulizations

### a. Comparing visualizations

### b. Designing an analysis

### c. Check any assumptions and run your analysis

### d. Create a visualization

### e. Write a caption

### f. Write about your results

### g. Sharing your affective visualization
